---
title: "Practical Machine Learning Course Project"
author: "Tony Dreher"
date: "Wednesday, June 10, 2015"
output: html_document
---

#Final Project Report - Practical Machine Learning Course

Tony Dreher


##Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Project Intended Results
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

##Reproduceablity
The following libraries were used for this project.  The corresponding packages may need to be installed before loading the libraries.

```{r,message=FALSE,warning=FALSE,results='hide'}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
```

The following seed will be used for the purposes of reproducability.
```{r}
set.seed(666)
```

##Data
A training dataset and testing dataset have been provided for this exercise.

```{r}
training <- read.csv(url("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"), na.strings=c("NA","#DIV/0!",""))
```


##Cleaning the data
The following transformations were used to clean the data:

The first seven columns will be removed as they are irrelevant to the study at hand.

```{r}
names(training)[1:7]
training <- training[,-c(1:7)]
testing <- testing[,-c(1:7)]
dim(training);dim(testing)
```

Variables which have more than a 60% threshold of NAs will also be removed.
```{r}
training <-training[,colSums(!is.na(training))/nrow(training) >= 0.6]
testing <-testing[,colSums(!is.na(testing))/nrow(testing) >= 0.6]
dim(training);dim(testing)
```

Variables which lack variability would usually be removed as well, but in this case, there are none.

```{r}
nzv <- nearZeroVar(training,saveMetrics = TRUE)
nzv[nzv$nzv == TRUE,]
```

##Partioning the training set into two

Because the training set is so large and the testing set so small, a training subset is created with 60% of the original training data set to be used for training while the other 40% is used as an additional testing set before using the original testing data.

```{r}
inTrain <- createDataPartition(training$classe, p = .60,list = FALSE)
trainingsubset <- training[ inTrain,]
testingsubset <- training[-inTrain,]
```

##Decision Tree Model

Since the a categorical outcome is being predicted, the decision tree model implemented first using the method `rpart`.

```{r}
modFit<-train(classe~.,method="rpart", data=trainingsubset)
print(modFit,digits=3)
print(modFit$finalModel)
fancyRpartPlot(modFit$finalModel,main="Decision Tree")
classepredict <- predict(modFit,testingsubset)
confusionMatrix(testingsubset$classe,classepredict)
```
Unfortunately, the decision tree model was not that accurate, 48% overall, and it was particularly bad at predicting outcome D. 

##Random Forest Model
Because the decision tree model was so inaccurate and the outcomes seem to be more nuanced than first suspected, a random forest model may be more appropriate.

```{r}
modFit2 <- train(classe ~ ., method="rf",trControl=trainControl(method = "cv", number = 4), data=trainingsubset)
print(modFit2,digits=3)
classepredict2 <- predict(modFit2,testingsubset)
confusionMatrix(testingsubset$classe,classepredict2)
```
The random forest model has a 99% accuracy - much better than the decision tree model. Additionally, the specificity and sensitivity is in the 90s across all 5 classes.


##In Sample & Out of Sample Error Rates
The in-sample error rate is 1% (1 - accuracy of the in-sample test).  The out-of-sample error rate is 1.01% (1 - accuracy of prediction model used on the test data).  

##Submission

```{r,eval=FALSE}
predictfinal <- predict(modFit2, newdata=testing)
predictfinal

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)

```
